---
title: "Modelling horizontal-facing benthic drop camera imagery"
author: "Claude Spencer"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Visualise the data

In this stage, we will load and visualise tidy data exported from CheckEM (https://marine-ecology.shinyapps.io/checkem).

Load libraries 
```{r, results='hide', warning=FALSE, message=FALSE}
library(tidyverse)
library(GlobalArchive)
library(ggplot2)
library(scatterpie)
library(sf)
library(ggnewscale)
library(terra)
library(stars)
library(starsExtra)
```

Set the study name
```{r}
name <- '2021-2022_SwC_BOSS'
```

Load metadata
```{r}
metadata <- read.csv(paste0("data/tidy/", name, "_Metadata.csv")) %>%
  glimpse()
```

Load habitat data and join with metadata
```{r}
tidy.habitat <- read.csv(paste0("data/tidy/", name, "_Habitat.csv")) %>%
  left_join(metadata) %>%
  dplyr::filter(!level_2 %in% "Unscorable") %>%
  # Make broad habitat levels for plotting
  dplyr::mutate(habitat = case_when(level_2 %in% c("Sponges", "Cnidaria",
                                                   "Bryozoa", "Sessile invertebrates",
                                                   "Ascidians", "Echinoderms") ~ "Sessile invertebrates",
                                    level_2 %in% "Macroalgae" ~ "Macroalgae",
                                    level_2 %in% "Seagrasses" ~ "Seagrass",
                                    level_2 %in% "Substrate" & level_3 %in% "Unconsolidated (soft)"~ "Sand",
                                    level_2 %in% "Substrate" & level_3 %in% "Consolidated (hard)"~ "Rock")) %>%
  pivot_wider(names_from = habitat, values_from = number, values_fill = 0) %>%
  glimpse()
```

Load shapefiles for plotting
```{r}
sf_use_s2(F)                                                                    # Needed so st_crop doesn't error when cropping with lat longs                                                                   # Otherwise st_crop errors - need to find a better way
# Marine park data from CAPAD 2022
marine.parks <- st_read("data/spatial/shapefiles/Collaborative_Australian_Protected_Areas_Database_(CAPAD)_2022_-_Marine.shp") %>%
  st_crop(xmin = min(tidy.habitat$longitude),
          xmax = max(tidy.habitat$longitude),
          ymin = min(tidy.habitat$latitude),
          ymax = max(tidy.habitat$latitude)) %>%
  dplyr::mutate(ZONE_TYPE = str_replace_all(ZONE_TYPE, 
                                            "\\s*\\([^\\)]+\\)", "")) %>%
  dplyr::filter(ZONE_TYPE %in% c("National Park Zone", "Sanctuary Zone"))
```

High res outline for Australia
```{r}
aus <- st_read("data/spatial/shapefiles/cstauscd_r.mif", crs = 4283) %>%
  dplyr::filter(!FEAT_CODE %in% "sea") %>%
  st_crop(xmin = min(tidy.habitat$longitude),
          xmax = max(tidy.habitat$longitude),
          ymin = min(tidy.habitat$latitude),
          ymax = max(tidy.habitat$latitude))
```

Coastal waters limit
```{r}
cwatr <- st_read("data/spatial/shapefiles/amb_coastal_waters_limit.shp") %>%
  st_crop(xmin = min(tidy.habitat$longitude),
          xmax = max(tidy.habitat$longitude),
          ymin = min(tidy.habitat$latitude),
          ymax = max(tidy.habitat$latitude))
```

```{r}
hab_fills <- scale_fill_manual(values = c("Sessile invertebrates" = "plum",
                             "Macroalgae" = "darkgoldenrod4",
                             "Seagrass" = "forestgreen",
                             "Rock" = "grey40",
                             "Sand" = "wheat"), 
                  name = "Habitat")
```

Visualise data as a scatterpie
```{r}
ggplot() +
  geom_sf(data = aus, fill = "seashell2", colour = "black", size = 0.2) +
  geom_sf(data = marine.parks, aes(fill = ZONE_TYPE), alpha = 2/5, colour = NA) +
  scale_fill_manual(values = c("National Park Zone" = "#7bbc63",
                               "Sanctuary Zone" = "#bfd054"),
                    name = "Marine Parks") +
  new_scale_fill() +
  geom_sf(data = cwatr, colour = "firebrick", alpha = 4/5, size = 0.3) +
  geom_scatterpie(data = tidy.habitat, aes(x = longitude, y = latitude),
                  cols = c("Sessile invertebrates", "Macroalgae", "Seagrass", "Rock", "Sand"),
                  pie_scale = 0.25, colour = NA) +
  hab_fills +
  labs(x = "Longitude", y = "Latitude") +
  coord_sf(xlim = c(114.75, 114.95), ylim = c(-34.01, -34.14)) +
  theme_minimal() +
  theme(panel.background = element_rect(fill = "#b9d1d6"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank())
```

## Load in spatial bathymetry data and extract bathymetry derivatives at each sample point

Load bathymetry data
```{r}
bathy <- rast("data/spatial/rasters/swc_multibeam_UTM50.tif")
```

Create bathymetry derivatives
```{r}
preds <- terrain(bathy, neighbors = 8,
                 v = c("slope", "aspect", "TPI", "TRI", "roughness"),
                 unit = "degrees")
```

Calculate detrended bathymetry
```{r}
zstar <- st_as_stars(bathy)                                                    
detre <- detrend(zstar, parallel = 8)                                           
detre <- as(object = detre, Class = "SpatRaster")                               
names(detre) <- c("detrended", "lineartrend")
```

Stack the derivatives
```{r}
preds <- rast(list(bathy, preds, detre[[1]]))                                   
names(preds)[1] <- "mbdepth"
```

Load metadata
```{r}
metadata <- read.csv(paste0("data/tidy/", name, "_Metadata.csv")) %>%
  glimpse()
```

Load habitat data and join with metadata
```{r}
tidy.habitat <- read.csv(paste0("data/tidy/", name, "_Habitat.csv")) %>%
  left_join(metadata) %>%
  dplyr::filter(!level_2 %in% "Unscorable") %>%
  # Make broad habitat levels for modelling
  dplyr::mutate(habitat = case_when(level_2 %in% c("Sponges", "Cnidaria",
                                                   "Bryozoa", "Sessile invertebrates",
                                                   "Ascidians", "Echinoderms") ~ "sessile.invertebrates",
                                    level_2 %in% "Macroalgae" ~ "macroalgae",
                                    level_2 %in% "Seagrasses" ~ "seagrasses",
                                    level_2 %in% "Substrate" & level_3 %in% "Unconsolidated (soft)"~ "sand",
                                    level_2 %in% "Substrate" & level_3 %in% "Consolidated (hard)"~ "rock")) %>%
  group_by(campaignid, sample) %>%
  dplyr::mutate(total.points.annotated = sum(number)) %>%
  ungroup() %>%
  glimpse()
```

Transform the habitat to a SpatVector
```{r}
habitat.vect <- vect(tidy.habitat, geom = c("longitude", "latitude"), crs = "epsg:4326") %>%
  project(preds)                                                                

plot(preds[[1]])
plot(habitat.vect, add = T)
```

Run the extraction process
```{r}
tidy.habitat_t   <- as.data.frame(habitat.vect, geom = "XY") %>%               
  left_join(tidy.habitat)                                                       

habitat.bathy.derivatives   <- cbind(tidy.habitat_t, 
                                     terra::extract(preds, habitat.vect)) %>%   
  dplyr::filter(!is.na(depth),                                                  
                !is.na(roughness)) %>%
  glimpse()
```

# Model selection
Bring in and format the data

```{r}
dat <- readRDS(paste0("data/tidy/", name, "_habitat-bathymetry.rds")) %>%
  dplyr::mutate(mbdepth = abs(mbdepth)) %>%                                     
  glimpse()
```

Set predictor variables
```{r}
names(dat)
pred.vars <- c("mbdepth","roughness", "detrended", 
               "slope", "TPI", "aspect", "TRI")  
```

Check for correlation of predictor variables
Remove anything highly correlated (>0.95)
```{r}
round(cor(dat[ , pred.vars]), 2)
```

Review of individual predictors for even distribution
Plot of likely transformations
```{r}
par(mfrow = c(3, 2))
for (i in pred.vars) {
  x <-dat[ , i]
  x = as.numeric(unlist(x))
  hist((x))
  plot((x), main = paste(i))
  hist(sqrt(x))
  plot(sqrt(x))
  hist(log(x + 1))
  plot(log(x + 1))
}
```

Check to make sure response variables have less than 80% zeros
```{r}
unique.vars = unique(as.character(dat$habitat))
unique.vars.use = character()
for(i in 1:length(unique.vars)){
  temp.dat = dat[which(dat$habitat == unique.vars[i]),]
  if(length(which(temp.dat$habitat == 0)) / nrow(temp.dat) < 0.8){
    unique.vars.use = c(unique.vars.use, unique.vars[i])}
}
unique.vars.use   
```

Set-up the environment to run model selection
```{r}
outdir    <- ("model out/")                                                     
use.dat   <- dat[dat$habitat %in% c(unique.vars.use), ]
out.all   <- list()
var.imp   <- list()
```

Re-set predictor and response variables
```{r}
pred.vars <- c("mbdepth","roughness", "detrended", 
               "TPI", "aspect")
resp.vars <- unique.vars.use
```

Run the full subset model selection process
```{r}
for(i in 1:length(resp.vars)){
  print(resp.vars[i])
  use.dat <- dat[dat$habitat == resp.vars[i],]
  use.dat   <- as.data.frame(use.dat)
  Model1  <- gam(cbind(number, (total.points.annotated - number)) ~ 
                   s(mbdepth, bs = 'cr'),
                 family = binomial("logit"),  data = use.dat)
  
  model.set <- generate.model.set(use.dat = use.dat,
                                  test.fit = Model1,
                                  pred.vars.cont = pred.vars,
                                  cyclic.vars = c("aspect"),
                                  k = 5,
                                  cov.cutoff = 0.7
  )
  out.list <- fit.model.set(model.set,
                            max.models = 600,
                            parallel = T)
  names(out.list)
  
  out.list$failed.models # examine the list of failed models
  mod.table <- out.list$mod.data.out  # look at the model selection table
  mod.table <- mod.table[order(mod.table$AICc), ]
  mod.table$cumsum.wi <- cumsum(mod.table$wi.AICc)
  out.i     <- mod.table[which(mod.table$delta.AICc <= 2), ]
  out.all   <- c(out.all, list(out.i))
  var.imp   <- c(var.imp, list(out.list$variable.importance$aic$variable.weights.raw))
  
  # Plot the best models
  for(m in 1:nrow(out.i)){
    best.model.name <- as.character(out.i$modname[m])
    
    png(file = paste(outdir, m, resp.vars[i], "mod_fits.png", sep = ""))
    if(best.model.name != "null"){
      par(mfrow = c(3, 1), mar = c(9, 4, 3, 1))
      best.model = out.list$success.models[[best.model.name]]
      plot(best.model, all.terms = T, pages = 1, residuals = T, pch = 16)
      mtext(side = 2, text = resp.vars[i], outer = F)}  
    dev.off()
  }
}
```

Save model fits and importance scores
```{r}
names(out.all) <- resp.vars
names(var.imp) <- resp.vars
all.mod.fits <- do.call("rbind", out.all)
all.var.imp  <- do.call("rbind", var.imp)
write.csv(all.mod.fits[ , -2], file = paste0(outdir, name, "_all.mod.fits.csv"))
write.csv(all.var.imp,         file = paste0(outdir, name, "_all.var.imp.csv"))
```

# Predict habitat
Load habitat data
```{r}
dat   <- readRDS(paste0("data/tidy/", name, "_Habitat-bathymetry.rds")) %>%
  dplyr::mutate(mbdepth = abs(mbdepth)) %>%                                     
  pivot_wider(names_from = habitat, values_from = number, values_fill = 0) %>%
  glimpse()
```

Load the bathy and derivatives
```{r}
preds  <- readRDS(paste0("data/spatial/rasters/", name, "_bathymetry-derivatives.rds"))
plot(preds)
```

Transform bathy to a dataframe to predict onto
```{r}
preddf <- as.data.frame(preds, xy = TRUE, na.rm = TRUE) %>%
  dplyr::mutate(mbdepth = abs(mbdepth))
```

Set models using the formula from top model
Sessile invertebrates
```{r}
m_inverts <- gam(cbind(sessile.invertebrates, total.points.annotated - sessile.invertebrates) ~ 
                 s(mbdepth,     k = 5, bs = "cr")  + 
                 s(roughness, k = 5, bs = "cr") +
                 s(TPI, k = 5, bs = "cr"), 
               data = dat, method = "REML", family = binomial("logit"))
summary(m_inverts)
plot(m_inverts, pages = 1, residuals = T)
```

Rock
```{r}
m_rock <- gam(cbind(rock, total.points.annotated - rock) ~ 
                   s(detrended,     k = 5, bs = "cr")  + 
                   s(mbdepth, k = 5, bs = "cr") + 
                   s(roughness, k = 5, bs = "cr"), 
                 data = dat, method = "REML", family = binomial("logit"))
summary(m_rock)
plot(m_rock, pages = 1, residuals = T)
```

Sand
```{r}
m_sand <- gam(cbind(sand, total.points.annotated - sand) ~ 
                s(detrended,     k = 5, bs = "cr")  + 
                s(mbdepth, k = 5, bs = "cr") +
                s(TPI, k = 5, bs = "cr"), 
              data = dat, method = "REML", family = binomial("logit"))
summary(m_sand)
plot(m_sand, pages = 1, residuals = T)
```

Seagrasses
```{r}
m_seagrass <- gam(cbind(seagrasses, total.points.annotated - seagrasses) ~ 
                s(detrended,     k = 5, bs = "cr")  + 
                s(mbdepth, k = 5, bs = "cr") + 
                s(roughness, k = 5, bs = "cr"), 
              data = dat, method = "REML", family = binomial("logit"))
summary(m_seagrass)
plot(m_seagrass, pages = 1, residuals = T)
```

Macroalgae
m_macro <- gam(cbind(macroalgae, total.points.annotated - macroalgae) ~ 
                 s(detrended,     k = 5, bs = "cr")  + 
                 s(mbdepth, k = 5, bs = "cr") +
                 s(TPI, k = 5, bs = "cr"), 
               data = dat, method = "REML", family = binomial("logit"))
summary(m_macro)
plot(m_macro, pages = 1, residuals = T)

# Predict, rasterise and plot habitat predictions ----
preddf <- cbind(preddf, 
                "pmacro" = predict(m_macro, preddf, type = "response"),
                "prock" = predict(m_rock, preddf, type = "response"),
                "psand" = predict(m_sand, preddf, type = "response"),
                "pseagrass" = predict(m_seagrass, preddf, type = "response"),
                "pinverts" = predict(m_inverts, preddf, type = "response")) %>%
  glimpse()

prasts <- rast(preddf %>% dplyr::select(x, y, pmacro, prock, psand, pseagrass, pinverts),
                        crs = crs(preds)) %>%
  aggregate(fact = 5, fun = "mean")                                             # Aggregate to speed up plots 
plot(prasts)
summary(prasts)

# Transform back to a dataframe for tidy plotting ----
preddf <- as.data.frame(prasts, xy = T, na.rm = T) %>%
  glimpse()

# Categorise by dominant habitat ----
preddf$dom_tag <- apply(preddf %>% dplyr::select(pmacro, prock, psand, pseagrass, pinverts), 1,
                         FUN = function(x){names(which.max(x))})
preddf$dom_tag <- sub('.', '', preddf$dom_tag)
head(preddf)

